A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
Working on Extractable Container Extensions
Intro

%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newminted{coq}{fontsize=\small}

@@

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorithm. This is given by the
\haskell{partition} and \haskell{quicksort} function below (written in Haskell
code):
While \haskell{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \haskell{partition}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

A pen-and-paper technique would be to show that the \haskell{partition} is a
\emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
type $A$. Here the term ``well-founded'' refers to the fact that the infinite
iteration of this coalgebra will produce a \emph{finite} binary tree with nodes
in $A$.

@@

\mvol{Need some transiiton here to build up to listing the goals and contributions of the paper. Also, ``axiom-free'' and ``obj.magic'' come out of nowhere here, so both need to be established earlier in the introduction.}
The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.
\end{enumerate}
In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}

>>>>

%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}

@@

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorith, written in OCaml below:
\begin{minted}{OCaml}
  let divide = function
  |  [] -> None
  |  x::xs -> Some (x, List.partition (fun y -> y < x) xs);;

  let rec quicksort xs = match divide xs with
    | None -> []
    | Some (x, (smaller, larger)) -> quicksort smaller @ (x::quicksort larger)
\end{minted}
While \ocaml{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

The main approach for implementing nonstructural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are coupled with
\emph{termination proofs}. Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}. A benefit of
this approach is that, extracting verified nonstructural recursive functions to
OCaml will erase the termination proofs, and so produce code that will be
closer to what a programmer may have written directly in OCaml. However,
reasoning about program equivalences requires dealing with such termination
proofs, and it is common practice to use custom reduction lemmas that will be
used extensively when proving properties  about the nonstructural recursive
functions.

A better approach for reasoning about program equivalences comes from
\emph{structured recursion schemes}, and their associated algebraic laws.  In
particular, \ocaml{quicksort} is an instance of a \emph{divide-and-conquer}
algorithm which, in the functional programming literature is known as a
hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
that every recursive program can be turned into an hylomorphism, which implies
that, ``\emph{every recursion scheme can be formalised as an hylomorphism}''.
Furthermore, the laws that are satisfied by hylomorphisms ease reasoning about
program equivalences, since they can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to (semi-)automatic
parallelisations~\cite{Gibbons96:Third}.

Despite the benefits of hylomorphisms for reasoning about program equivalences,
not many approaches mechanise structured recursion schemes in proof assistants.
A recent approach by Abreu et al.\~\cite{AbreuDHJMS23} encode an algebraic
approach to divide-and-conquer computations, in which termination is entirely
enforced by the typing discipline of the recursion scheme. However, although
their approach solves the problem of termination proofs, as well as the
performance of the code that is run \emph{within Coq}, it no longer allows the
extraction of clean, naturally-looking code.  By naturally-looking and clean,
we mean code that (1) preserves the recursive structure intended by the
programmer; and (2) does not lead to unsafe casts \ocaml{Obj.magic} in the
generated code.

In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{quicksort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function
  | [] -> []
  | h :: t ->
    let (l, r) = partition (fun x0 -> (<=) x0 h) t in
    let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
    app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}

The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a strong focus on proving their algebraic laws and 
    program equivalences, and clean code extraction.
  \item We formalise algorithms for divide and conquer, dynamic programming and
    mutual recursion.
  \item We verified the well-known short-cut fusion optimisation in Coq and
    extract the resulting code to OCaml.
\end{itemize}

\dcas{Rewritten up to here}
In
\ocaml{quicksort} above, a pen-and-paper technique to prove termination would
be to show that \ocaml{divide} is a \emph{well-founded coalgebra} on the
functor $(1 + A \times - \times -)$ for a fixed type $A$. Here the term
``well-founded'' refers to the fact that the infinite iteration of this
coalgebra will produce a \emph{finite} binary tree with nodes in $A$.

@@

\mvol{Need some transiiton here to build up to listing the goals and contributions of the paper. Also, ``axiom-free'' and ``obj.magic'' come out of nowhere here, so both need to be established earlier in the introduction.}
The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \ocaml{Obj.magic}.
\end{enumerate}
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}
